---
title: "Run 1 simulation"

---

```{r}
library(ggplot2)
library(reshape2)
library(dplyr)
library(ggpubr)
library(igraph)
library(EnvStats)
library(aricode)
library(vegan)
library(ade4)
library(ape)
library(tidyr)
library(ggraph)

source(here::here("code/functions/functions_simulation.R")) 
source(here::here("code/functions/functions_simulation_inspection.R"))
source(here::here("code/functions/function_simulation_7may.R"))

```


Parameters for plotting

```{r}
my.theme<-theme(axis.text=element_text(size=12),
                axis.title = element_text(size = 14),
                legend.position = "top",
                legend.text=element_text(size=10),
                legend.title = element_text(size=12),
                plot.title = element_text(face="bold",size=14,margin=margin(0,0,20,0),hjust = 0.5),
                axis.title.y = element_text(margin = margin(t = 0, r = 15, b = 0, l = 0)),
                axis.title.x = element_text(margin = margin(t = 15, r = 0, b = 0, l = 0)))
```




## Parameters foodweb

```{r}
pars = list()

pars$competition_coefficient = 0.025
pars$Sbasal = 5 # number of basal species
pars$Smax = 1000 # Maximal number of species allowed
#pars$Bspe = 4 # Basal species impact on the speciation probality
#pars$Bext = 4 # Basal species impact on extinction probability
pars$av_r = 0.18#0.185 # 0.2 range of the niche
pars$sd = 0.5*pars$av_r + 0.0001 # Standard deviation of the normal distribution used to calculate the niche optimum trait of a new species
#pars$sd = 0.5*pars$av_r + 0.0001 # Standard deviation of the normal distribution used to calculate the niche optimum trait of a new species

# PROBABILITY OF MUTATION

pars$u_max = 0.23#0.15 #0.15 # mutation probability (0.075, )
pars$d = 0.5 # Decrease speed of the establishment probability
pars$I_max = 60 # Maximal number of interactioning species
pars$beta_n =  1 # parameter of the beta distribution 

# STRENGTH OF NEUTRAL-DRIVEN EVOLUTION

pars$SN = 0 # strength for neutral-driven evolution

# PROBABILITY OF ESTABLISHMENT

# strength of selection-driven selection is 1 - SN
pars$estab_prob_neutral = 0.5 # neutral probability of establishment

# Facilitation & Foodweb

pars$u_0pos = 1  
pars$u_1pos = -1 
pars$a_upos = 0.45 

# PROBABILITY OF EXTINCTION

pars$ext_prob_neutral = rep(0.0005, pars$Smax) # neutral probability of extinction

# Competition

pars$e_0neg = 0.1 #0.15 # Asymptotic extinction probability with infinite negative interactions
pars$a_eneg = 0.025 # Shape of the exponential decay of the negative extinction - interaction relationship
pars$e_1neg = -pars$e_0neg  # Extinction probability with absence of interactions

# Facilitation & Foodweb

pars$e_0pos = 0.075 
pars$e_1pos = 5.19 
pars$a_epos = 1.2 


#########################################

# Logistic function parameters
pars$k <- 10  # Steepness of the logistic curve
pars$midpoint <- 0.5  # Midpoint for the logistic curve

```


# parameters simulation

```{r}
nsteps =  150
nsim = 1

```




--------------------------
beta_ext * 0.097 + (1 - beta_ext) * avg_similarity

# New pars based on testing (13 sept)

```{r}
pars$u_0pos <- 0.4   # Baseline probability
pars$u_1pos <- 0.7  # Scaling factor
pars$a_upos <- 0.2   # Shape of the exponential increase

pars$e_0neg <- 0.01   # Baseline probability
pars$e_1neg <- 0.05 #0.5 # Scaling factor for out-degree
pars$a_eneg <- 0.2 #0.2  Shape of the exponential increase

pars$beta_ext <- 0.8 # Control strength between out-degree and similarity

pars$competition_coefficient <- 0.01 #0.1 #0.1 #0.06

pars$multiplier = 0.9 #1

```

```{r}
pars$beta_n =  50#50
```


Top predators with:


```{r}
pars$av_r = 0.17 #0.16
pars$beta_ext <- 0.8 # Control strength between out-degree and similarity
pars$competition_coefficient <- 0.0085#0.0085 #0.0025 #0.1 #0.1 #0.06
pars$multiplier = 1 # or also 0.8
```


```{r}
pars$beta_ext <- 0.6
```


```{r}
#saveRDS(pars, here::here("code/simulation_pars/pars_16May.rds"))
```


Making selection-based scenario more pronounced

```{r}
pars$u_0pos <- 0.2
pars$u_1pos <- 1.2
pars$a_upos <- 0.3
pars$e_0neg <- 0.03
pars$e_1neg <- 0.15
pars$a_eneg <- 0.6
pars$competition_coefficient <- 0.025
pars$ext_prob_neutral <- 0.0002
pars$beta_ext <- 0.5

#saveRDS(pars, "simulation_pars/pars_selection_17May.rds")
```


```{r}
pars$u_0pos <- 0.2
pars$u_1pos <- 1.2
pars$a_upos <- 0.3
pars$e_0neg <- 0.1#0.1
pars$e_1neg <- 0.3#0.6
pars$a_eneg <- 0.6#0.2
pars$competition_coefficient <- 0.2#0.2
pars$ext_prob_neutral <- 0.0002
pars$beta_ext <- 0.5
#pars$multiplier<- 2

#saveRDS(pars, "simulation_pars/pars_selection_20May.rds")
```


Trying to increase niche space

```{r}
pars$sd <- 0.15              # More trait exploration
# pars$a_upos <- 0.1           # Less steep establishment penalty
# pars$e_1neg <- 0.1           # Less top-down extinction pressure
# pars$competition_coefficient <- 0.1  # Less extinction from similarity

#pars$beta_ext <- 0.4

```


Tryng to slow S builtup speed

```{r}
# upper limit of speciation probability
# pars$u_max <- 0.2  # from 0.23 
# 
# # steepness of decline in speciation with diversity
# 
# pars$d <- 0.6  # from 0.5
# 
# pars$u_0pos <- 0.2
# pars$u_1pos <- 1.2
# pars$a_upos <- 0.3
# pars$e_0neg <- 0.05#0.1
# pars$e_1neg <- 0.1#0.6
# pars$a_eneg <- 0.2#0.2
# pars$competition_coefficient <- 0.1#0.2

```


Neutral

```{r}
# pars$SN = 1
# pars$ext_prob_neutral = 0.05
# pars$estab_prob_neutral = 0.5
```

```{r}
nsteps =  150
```


```{r}
#saveRDS(pars, "simulation_pars/pars_selection_20May.rds")
```


### Simulation


```{r}

list_res <- list()
seed_list <- NULL
seed_record <- rep(NA, nsim)
successful_sim <- 0
total_nb_sim <- 0

while (successful_sim < nsim) {
  seed <- sample.int(80000, 1)
  if (seed %in% seed_list) next
  seed_list <- c(seed_list, seed)
  
  cat("Trying seed:", seed, "\n")
  simulation <- sim_model(seed = seed, pars = pars, nsteps = nsteps)
  
  # ❌ If simulation failed (returned NULL), skip this iteration
  if (is.null(simulation)) {
    cat("❌ Simulation discarded (didn't meet criteria).\n")
    next
  }
  
  # ✅ Only store if simulation passed all criteria
  successful_sim <- successful_sim + 1
  total_nb_sim <- total_nb_sim + 1
  seed_record[successful_sim] <- seed
  
  res_sim <- list(
    simulation_number = paste0("simulation", successful_sim),
    seed = seed,
    parameters = pars,
    presence_matrix = simulation$pres,
    traits_df = simulation$traits,
    extinxction_matrix = simulation$extinct, 
    network_list = simulation$L_list,
    network_crop_list = simulation$L_cropped_list,
    anc_dist_table = simulation$dist_anc, 
    list_anc_dist = simulation$list_dist_anc
  )
  
  list_res[[successful_sim]] <- res_sim
  cat("✅ Stored simulation", successful_sim, "of", nsim, "\n")
}



```


```{r}
# Convert the list to a data.frame
ext_df <- do.call(rbind, lapply(seq_along(simulation$ext_prob_log), function(i) {
  data.frame(
    timestep = i * 20,
    ext_prob = simulation$ext_prob_log[[i]]
  )
}))

ggplot(ext_df, aes(x = ext_prob)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~ timestep, scales = "free_y") +
  labs(
    x = "Extinction probability (selection component)",
    y = "Count",
    title = "Extinction probability distribution across timesteps"
  ) +
  theme_minimal()

```



```{r}
#saveRDS(res_sim, here::here("output/temporary_output/res_sim.rds"))

# res_sim <- readRDS(here::here("output/testing/res_sim.rds"))


```



# Inspection

## Inspection of individual simulation

```{r}
inspect_simulation_fw(simulation_data = res_sim, nbasals = pars$Sbasal, Smax = pars$Smax)
```


## inspection of network structures

```{r}

#saveRDS(res_sim, here::here("output/results_toplot_networks_timesteps.rds"))
#res_sim <- readRDS(here::here("output/results_toplot_networks_timesteps.rds"))

plot_network_at_timestep_stratified <- function(matrix, timestep) {
  # Create an igraph object from the adjacency matrix
  g <- graph_from_adjacency_matrix(as.matrix(matrix), mode = "directed")
  
  # Identify species roles
  in_degree <- igraph::degree(g, mode = "in")   # Number of incoming edges (eaten by others)
  out_degree <- igraph::degree(g, mode = "out") # Number of outgoing edges (eat others)
  
  # Assign roles
  V(g)$role <- ifelse(in_degree > 0 & out_degree == 0, "Herbivore", 
                      ifelse(in_degree > 0 & out_degree > 0, "Intermediate", 
                             "Top Predator"))
  
  # Define y-coordinates for stratification based on species role
  V(g)$y <- ifelse(V(g)$role == "Herbivore", 1, 
                   ifelse(V(g)$role == "Intermediate", 2, 3))
  
  # Generate a custom layout with stratified y-coordinates and evenly spaced x-coordinates
  layout_df <- data.frame(x = 1:vcount(g), y = V(g)$y)
  
  # Create a ggraph plot with a custom layout
  plot <- ggraph(g, layout = layout_df) + 
    geom_edge_link(edge_width = 0.1, arrow = arrow(length = unit(2, 'mm')), end_cap = circle(3, 'mm')) + 
    geom_node_point(aes(color = role), size = 2) + # Plot nodes with color based on role
    scale_color_manual(values = c("Herbivore" = "#009E73", "Intermediate" = "#E69F00", "Top Predator" = "#CC79A7")) +
    labs(title = paste("Timestep", timestep), color = "Species Role") +
    theme_void() + # Use a minimal theme
    theme(
      legend.position = "bottom", 
      legend.text = element_text(size = 12),     # Increase legend text size
      legend.title = element_text(size = 14),    # Increase legend title size
      legend.key.size = unit(1.5, "cm")           # Increase legend key size
    )
  
  return(plot)
}


# Timesteps to plot
timesteps_to_plot <- c(20, 30, 40, 50:60, 70, 80, 90:100, 110, 120)


list_network_cropped <- res_sim$network_crop_list

 # list_network_cropped <- lapply(list_network, function(mat) {
 #   mat[-c(1:pars$Sbasal), ]  # Remove the first 25 rows
 # })

# Create plots for the specified timesteps using ggraph with stratified layout
plot_list <- lapply(timesteps_to_plot, function(timestep) {
  plot_network_at_timestep_stratified(list_network_cropped[[timestep]], timestep)
})

# Arrange the plots using ggarrange
final_plot <- ggarrange(plotlist = plot_list, ncol = 3, nrow = 3, common.legend = TRUE)

# Display the final plot
final_plot
```